Event History Analyses for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2024-10-20 manuscript/manuscript.Rmd
update ms.

1a. Wrangle raw data and descriptive stats for one independent variable.
1b. Wrangle raw data and descriptive stats for two independent variables.
2a. Bayesian multilevel modeling for h(t)
2b. Bayesian multilevel modeling for ca(t)
3a. Frequentist multilevel modeling for h(t)
3b. Frequentist multilevel modeling for ca(t)

@@

Models for discrete-time h(t) and ca(t) can quickly become very complex when adding more than 1 time scale, due to many higher-order interactions. 
For example, model M4 contains two time scales as covariates: the passage of time on the across-bin or within-trial time scale (variable period_9), and the passage of time on the across-trial or within-experiment time scale (variable trial_c). When trials are presented in blocks, and blocks of trials within sessions, and when the experiment comprises three sessions, then four time scales can be defined (across-bin or within-trial, across-trial or within-block, across-block or within-session, and across-session or within-experiment).
From a theoretical perspective, adding more than 1 time scale is important to capture plasticity (e.g., proactive control) and other learning effects that play out on such longer time scales (across-trials, across-blocks, across-sessions), and that are probably present in each experiment in general.
From a practical perspective, therefore, it might be interesting for interpretational purposes to limit the number of experimental variables, because adding time scales quickly increases model complexity.

@@

Another approach: fit models to individual subjects and describe prevalence...

@@

We did not discuss continuous-time hazard analysis. As indicated by --allison -- learning discrete-time methods first will help in learning continuous-time methods. Given that RT is typically treated as a continuous variable, it is possible that continuous-time methods will ultimately prevail. However, they require much more data to estimate the continuous-time hazard (rate) function well. Thus, by trading a bit of temporal resolution for a lower number of trials, discrete-time methods seem ideal for dealing with typical psychological time-to-event data sets for which there are less than ~200 trials per condition per experiment.

@@

The hazard models in this tutorial assume that there is one event of interest. For RT data, this event constitutes a single transition between an "idle" state and a "responded" state. However, in certain situations, more than one event of interest might exist. For example, in a medical or health-related context, an individual might transition back and forth between a "healthy" state and a "depressed" state, before a final "death" state. When you have data on the timing of these transitions, one can apply multi-state models, which generalize survival analysis to transitions between three or more states [@SteeleF2004].
Also, the predictor variables in this tutorial are time-invariant, i.e., their value did not change over the course of a trial. Thus, another extension is to include time-varying predictors, i.e., predictors whose value can change across the time bins within a trial (REF). [[give a concrete example for this latter point]]

>>>>

* 1a. Wrangle raw data and descriptive stats for one independent variable.
* 1b. Wrangle raw data and descriptive stats for two independent variables.
* 2a. Bayesian multilevel modeling for h(t)
* 2b. Bayesian multilevel modeling for ca(t)
* 3a. Frequentist multilevel modeling for h(t)
* 3b. Frequentist multilevel modeling for ca(t)

@@

Models for discrete-time h(t) and ca(t) can quickly become very complex when adding more than 1 time scale, due to many higher-order interactions. 
For example, model M4 contains two time scales as covariates: the passage of time on the across-bin or within-trial time scale (variable period_9), and the passage of time on the across-trial or within-experiment time scale (variable trial_c). However, when trials are presented in blocks, and blocks of trials within sessions, and when the experiment comprises three sessions, then four time scales can be defined (across-bin or within-trial, across-trial or within-block, across-block or within-session, and across-session or within-experiment).
From a theoretical perspective, adding more than 1 time scale is important to capture plasticity (e.g., proactive control) and other learning effects that play out on such longer time scales (across-trials, across-blocks, across-sessions), and that are probably present in each experiment in general.
From a practical perspective, therefore, it might be interesting for interpretational purposes to limit the number of experimental variables, because adding time scales quickly increases model complexity.

@@

Another approach: fit models to individual subjects and describe prevalence...REF

@@

We did not discuss continuous-time hazard analysis. As indicated by @allisonSurvivalAnalysisUsing2010, learning discrete-time methods first will help in learning continuous-time methods. Given that RT is typically treated as a continuous variable, it is possible that continuous-time methods will ultimately prevail. However, they require much more data to estimate the continuous-time hazard (rate) function well. Thus, by trading a bit of temporal resolution for a lower number of trials, discrete-time methods seem ideal for dealing with typical psychological time-to-event data sets for which there are less than ~200 trials per condition per experiment.

@@

The hazard models in this tutorial assume that there is one event of interest. For RT data, this event constitutes a single transition between an "idle" state and a "responded" state. However, in certain situations, more than one event of interest might exist. For example, in a medical or health-related context, an individual might transition back and forth between a "healthy" state and a "depressed" state, before a final "death" state. When you have data on the timing of these transitions, one can apply multi-state models, which generalize survival analysis to transitions between three or more states [@SteeleF2004].
Also, the predictor variables in this tutorial are time-invariant, i.e., their value did not change over the course of a trial. Thus, another extension is to include time-varying predictors, i.e., predictors whose value can change across the time bins within a trial [@@allisonSurvivalAnalysisUsing2010]. For example, when gaze position is tracked during a visual search trial, the gaze-target distance will vary during a trial when the eyes move around before a manual response is given; shorter gaze-target distances should be associated with a higher hazard of response occurrence.

### 2024-10-20 manuscript/manuscript.tex
update ms.

In Tutorial 1a, we illustrate how to process or ``wrangle'' a previously published RT + accuracy data set to calculate descriptive statistics when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 1b we provide a generalisation of this approach to illustrate how one might descriptive statistics when using a more complex design, such as when there are two independent variables. In Tutorial 2a, we illustrate how one can fit Bayesian multi-level regression models to RT-hazard data using the R package brms. We discuss possible link functions, and plot the model-based effects of our predictors of interest. In Tutorial 2b we fit Bayesian multi-level regression models to \emph{timed} accuracy data to perform a micro-level speed-accuracy tradeoff (SAT) analysis, which complements the hazard analysis. In Tutorial 3a, we illustrate how to fit the same type of hazard regression models in a frequentist framework using the R package lme4. We then briefly compare and contrast these inferential frameworks when applied to EHA. In Tutorial 3b, we illustrate how to perform the SAT analysis in a frequentist framework.

@@

In the context of experimental psychology, it is common for participants to be presented with a task that has a right and a wrong answer. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time hazard function can be extended with the discrete-time conditional accuracy function (see equation 5 in part A of the supplementary material), which gives you the probability that a response is correct given that it is emitted in time bin t (Allison, 2010; Kantowitz \& Pachella, 2021; Wickelgren, 1977).

@@

Tutorials 1a and 1b show how to calculate and plot the descriptive statistics when there are one and two independent variables, respectively. Tutorials 2a and 2b illustrate how to use Bayesian multilevel modeling to fit hazard and conditional accuracy models, respectively. Tutorials 3a and 3b show how to implement, respectively, multilevel models for hazard and conditional accuracy in the frequentist framework.
Additionally, to further simplify the process for other users, the tutorials rely on a set of our own custom functions that make sub-processes easier to autmate, such as data wrangling and plotting functions (see part B in the supplemental material for a list of the custom functions).

@@

Generalisation (T4). Should this be online in Supps?? It would make the main text shorter and simpler, but make it no less available. We could just have a sentence at the end of T1, which says that we provide a generalisation and extension in T4, which is in Supps.

Plannng (T5) - if we get a simulation and power analysis script working, which we are happy with then we could include it here.

@@

In this third tutorial, we illustrate how to fit Bayesian hazard regression models to the masked response priming data set used in the first tutorial. Fitting (Bayesian or non-Bayesian) regression models to the data is important when you want to study how the shape of the hazard function depends on various predictors (Singer \& Willett, 2003).

@@

In this tutorial we will fit four Bayesian multilevel models (i.e., generalized linear mixed models) to the person-trial-bin oriented data set that we created in Tutorial 1a. We select the analysis range (200,600{]} and the cloglog link. The data is prepared as follows.

@@

To get the posterior distribution of each model parameter given the data, we need to specify a prior distribution for each parameter. The middle column of Figure 9 in part E of the supplementary material shows seven examples of prior distributions on the logit and/or cloglog scales.

While a normal distribution with relatively large variance is often used as a weakly informative prior for continuous dependent variables, rows A and B in Figure 9 show that specifying such distributions on the logit and cloglog scales leads to rather informative distributions on the original probability scale, as most mass is pushed to probabilities of 0 and 1. The other rows in Figure 9 show prior distributions on the logit and cloglog scale that we use instead.

@@

When the shape of the hazard function is rather smooth, one can fit a more parsemoneous model by using a polynomial specification of TIME. For our second example model, we thus use a third-order polynomial specification of TIME for the baseline condition (blank prime), and again assume that the effects of prime-target congruency and centered trial number are proportional and additive, and that the effect of trial number is linear. The model formula for model M2 looks as follows.

@@

So far, we assumed that the effect of our predictors condition and centered trial number are the same in each time bin. However, the descriptive plots suggest that the effect of condition can vary across time bins.
For the third model, we thus use a third-order polynomial specification of TIME for the baseline condition (blank prime), and relax the proportionality assumption for the predictor variables prime-target congruency (variable ``condition'') and trial number (variable ``trial\_c'').

@@

Figure 6 shows the point and effects of congruent and incongruent primes relative to neutral primes, for each time bin in trial numbers 500, 1000, and 1500.

@@

\caption{50/80/95 percentile intervals of the draws from the posterior distributions representing the effect of congruent and incongruent primes relative to neutral primes in trial number 1000.}\label{fig:plot-prime-effects}
\end{figure}

Table 4 shows the summaries of the draws from the posterior distributions of the effects of congruent and incongurent primes relative to the blank prime condition in trials 500, 1000, and 1500, in terms of a point estimate (the mean) and the upper and lower limits of the 95\% highest density interval. Also, by exponentiating the mean we obtain an effect size in terms of a hazard ratio.

@@

Figure 8 shows the point and effects of congruent and incongruent primes relative to neutral primes, for each time bin in trial numbers 500, 1000, and 1500.

@@

\caption{50/80/95 percentile intervals of the draws from the posterior distributions representing the effect of congruent and incongruent primes relative to neutral primes in trial numbers 500, 1000, and 1500.}\label{fig:plot-prime-ca-effects}
\end{figure}

@@

This main motivation for writing this paper is the observation that event history analysis remains under-used in psychological research, which means the field of research is not taking full advantage of the many benefits EHA provides compared to more conventional analyses. By providing a freely available set of tutorials, which provide step-by-step guidelines and ready-to-use R code, we hope that researchers will feel more comfortable using EHA in the future. Indeed, we hope that our tutorials may help to overcome a barrier to entry with EHA, which is the increase in analytical complexity compared to mean-average comparisons. While we have focused here on within-subject, factorial, small-\emph{N} designs, it is important to realize that event history analysis can be applied to other designs as well (large-\emph{N} designs with only one measurement per subject, between-subject designs, etc.). As such, the general workflow and associated code can be modified and applied more broadly to other contexts and research questions. In the following, we discuss issues relating to individual differences, limitations of the approach, and future extensions.

@@

\subsection{5.2 Individual differences}\label{individual-differences}

@@

\subsection{5.3 Limitation(s)}\label{limitations}

@@

We did not discuss continuous-time hazard analysis. As indicated by --allison -- learning discrete-time methods first will help in learning continuous-time methods. Given that RT is typically treated as a continuous variable, it is possible that continuous-time methods will ultimately prevail. However, they require much more data to estimate the continuous-time hazard (rate) function well. Thus, by trading a bit of temporal resolution for a lower number of trials, discrete-time methods seem ideal for dealing with typical psychological time-to-event data sets for which there are less than \textasciitilde200 trials per condition per experiment.

\subsection{5.4 Extensions}\label{extensions}

The hazard models in this tutorial assume that there is one event of interest. For RT data, this event constitutes a single transition between an ``idle'' state and a ``responded'' state. However, in certain situations, more than one event of interest might exist. For example, in a medical or health-related context, an individual might transition back and forth between a ``healthy'' state and a ``depressed'' state, before a final ``death'' state. When you have data on the timing of these transitions, one can apply multi-state models, which generalize survival analysis to transitions between three or more states (Steele, Goldstein, \& Browne, 2004).
Also, the predictor variables in this tutorial are time-invariant, i.e., their value did not change over the course of a trial. Thus, another extension is to include time-varying predictors, i.e., predictors whose value can change across the time bins within a trial (REF). {[}{[}give a concrete example for this latter point{]}{]}

>>>>

In Tutorial 1a, we illustrate how to process or ``wrangle'' a previously published RT + accuracy data set to calculate descriptive statistics when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 1b we provide a generalisation of this approach to illustrate how one can calculate the descriptive statistics when using a more complex design, such as when there are two independent variables. In Tutorial 2a, we illustrate how one can fit Bayesian multi-level regression models to RT data using the R package brms. We discuss possible link functions, and plot the model-based effects of our predictors of interest. In Tutorial 2b we fit Bayesian multi-level regression models to \emph{timed} accuracy data to perform a micro-level speed-accuracy tradeoff (SAT) analysis, which complements the hazard analysis. In Tutorial 3a, we illustrate how to fit the same type of hazard regression models in a frequentist framework using the R package lme4. We then briefly compare and contrast these inferential frameworks when applied to EHA. In Tutorial 3b, we illustrate how to perform the SAT analysis in a frequentist framework.

@@

In the context of experimental psychology, it is common for participants to be presented with a task that has a right and a wrong answer. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time hazard analysis can be extended with a discrete-time SAT analyis. Specifically, the hazard function of event occurrence can be extended with the discrete-time conditional accuracy function (see equation 5 in part A of the supplementary material), which gives you the probability that a response is correct given that it is emitted in time bin t (Allison, 2010; Kantowitz \& Pachella, 2021; Wickelgren, 1977).

@@

Tutorials 1a and 1b show how to calculate and plot the descriptive statistics when there are one and two independent variables, respectively. Tutorials 2a and 2b illustrate how to use Bayesian multilevel modeling to fit hazard and conditional accuracy models, respectively. Tutorials 3a and 3b show how to implement, respectively, multilevel models for hazard and conditional accuracy in the frequentist framework.
Additionally, to further simplify the process for other users, the tutorials rely on a set of our own custom functions that make sub-processes easier to automate, such as data wrangling and plotting functions (see part B in the supplemental material for a list of the custom functions).

@@

Plannng (T4) - if we get a simulation and power analysis script working, which we are happy with then we could include it here.

@@

In this third tutorial, we illustrate how to fit Bayesian hazard regression models to the masked response priming data set used in Tutorial 1a. Fitting (Bayesian or non-Bayesian) regression models to the data is important when you want to study how the shape of the hazard function depends on various predictors (Singer \& Willett, 2003).

@@

In this tutorial we will fit four Bayesian multilevel models (i.e., generalized linear mixed models) that differ in complexity to the person-trial-bin oriented data set that we created in Tutorial 1a. We select the analysis range (200,600{]} and the cloglog link. The data is prepared as follows.

@@

To get the posterior distribution of each model parameter given the data, we need to specify a prior distribution for each parameter. The middle column of Figure 12 in part E of the supplementary material shows seven examples of prior distributions on the logit and/or cloglog scales.

While a normal distribution with relatively large variance is often used as a weakly informative prior for continuous dependent variables, rows A and B in Figure 12 show that specifying such distributions on the logit and cloglog scales leads to rather informative distributions on the original probability scale, as most mass is pushed to probabilities of 0 and 1. The other rows in Figure 9 show prior distributions on the logit and cloglog scale that we use instead.

@@

When the shape of the hazard function is rather smooth, one can fit a more parsimonious model by using a polynomial specification of TIME. For our second example model, we thus use a third-order polynomial specification of TIME for the baseline condition (blank prime), and again assume that the effects of prime-target congruency and centered trial number are proportional and additive, and that the effect of trial number is linear. The model formula for model M2 looks as follows.

@@

So far, we assumed that the effect of our predictors condition and centered trial number are the same in each time bin. However, the descriptive plots suggest that the effect of condition can vary across time bins.
For the third model, we thus use a third-order polynomial specification of TIME for the baseline condition (blank prime), and relax the proportionality assumption for the predictor variables prime-target congruency (variable ``condition'') and centered trial number (variable ``trial\_c'').

@@

Figure 6 shows one point (mean) and three highest density interval (50/80/95\%) estimates for the effects of congruent and incongruent primes relative to neutral primes, for each time bin in trial numbers 500, 1000, and 1500.

@@

\caption{Means and 50/80/95\% highest density intervals of the draws from the posterior distributions representing the effect of congruent and incongruent primes relative to neutral primes in each bin for trial numbers 500, 1000, and 1500.}\label{fig:plot-prime-effects}
\end{figure}

Table 4 shows the summaries of the draws from the posterior distributions of the effects of congruent and incongurent primes relative to the blank prime condition in trials 500, 1000, and 1500, in terms of a point estimate (the mean) and the upper and lower bounds of the 95\% highest density interval. Also, by exponentiating the mean we obtain an effect size in terms of a hazard ratio.

@@

Figure 8 shows one point (mean) and three highest density interval (50/80/95\%) estimates for the effects of congruent and incongruent primes relative to neutral primes on logit-ca, for each time bin in trial numbers 500, 1000, and 1500.

@@

\caption{Means and 50/80/95\% highest density intervals of the draws from the posterior distributions representing the effect of congruent and incongruent primes relative to neutral primes in each bin for trial numbers 500, 1000, and 1500.}\label{fig:plot-prime-ca-effects}
\end{figure}

@@

This main motivation for writing this paper is the observation that event history analysis remains under-used in psychological research, which means the field of research is not taking full advantage of the many benefits EHA provides compared to more conventional analyses. By providing a freely available set of tutorials, which provide step-by-step guidelines and ready-to-use R code, we hope that researchers will feel more comfortable using EHA in the future. Indeed, we hope that our tutorials may help to overcome a barrier to entry with EHA, which is the increase in analytical complexity compared to mean-average comparisons. While we have focused here on within-subject, factorial, small-\emph{N} designs, it is important to realize that event history analysis can be applied to other designs as well (large-\emph{N} designs with only one measurement per subject, between-subject designs, etc.). As such, the general workflow and associated code can be modified and applied more broadly to other contexts and research questions. In the following, we discuss issues relating to model complexity versus interpretability, individual differences, limitations of the approach, and future extensions.

@@

\subsection{5.2 Model complexity versus interpretability}\label{model-complexity-versus-interpretability}

Models for discrete-time h(t) and ca(t) can quickly become very complex when adding more than 1 time scale, due to many higher-order interactions.
For example, model M4 contains two time scales as covariates: the passage of time on the across-bin or within-trial time scale (variable period\_9), and the passage of time on the across-trial or within-experiment time scale (variable trial\_c). However, when trials are presented in blocks, and blocks of trials within sessions, and when the experiment comprises three sessions, then four time scales can be defined (across-bin or within-trial, across-trial or within-block, across-block or within-session, and across-session or within-experiment).
From a theoretical perspective, adding more than 1 time scale is important to capture plasticity (e.g., proactive control) and other learning effects that play out on such longer time scales (across-trials, across-blocks, across-sessions), and that are probably present in each experiment in general.
From a practical perspective, therefore, it might be interesting for interpretational purposes to limit the number of experimental variables, because adding time scales quickly increases model complexity.

\subsection{5.3 Individual differences}\label{individual-differences}

@@

Another approach: fit models to individual subjects and describe prevalence\ldots REF

\subsection{5.4 Limitation(s)}\label{limitations}

@@

We did not discuss continuous-time hazard analysis. As indicated by Allison (2010), learning discrete-time methods first will help in learning continuous-time methods. Given that RT is typically treated as a continuous variable, it is possible that continuous-time methods will ultimately prevail. However, they require much more data to estimate the continuous-time hazard (rate) function well. Thus, by trading a bit of temporal resolution for a lower number of trials, discrete-time methods seem ideal for dealing with typical psychological time-to-event data sets for which there are less than \textasciitilde200 trials per condition per experiment.

\subsection{5.5 Extensions}\label{extensions}

The hazard models in this tutorial assume that there is one event of interest. For RT data, this event constitutes a single transition between an ``idle'' state and a ``responded'' state. However, in certain situations, more than one event of interest might exist. For example, in a medical or health-related context, an individual might transition back and forth between a ``healthy'' state and a ``depressed'' state, before a final ``death'' state. When you have data on the timing of these transitions, one can apply multi-state models, which generalize survival analysis to transitions between three or more states (Steele, Goldstein, \& Browne, 2004).
Also, the predictor variables in this tutorial are time-invariant, i.e., their value did not change over the course of a trial. Thus, another extension is to include time-varying predictors, i.e., predictors whose value can change across the time bins within a trial (@ Allison, 2010). For example, when gaze position is tracked during a visual search trial, the gaze-target distance will vary during a trial when the eyes move around before a manual response is given; shorter gaze-target distances should be associated with a higher hazard of response occurrence.
